\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-12-04 chapters/bci.tex
Proofread preface

Some impressive recent breakthroughs in speech and motor \ac{bci} for communication have
been realized using invasive microelectrode arrays~\cite{Willett2021} and
$\mu$\Ac{ecog}~\cite{Metzger2023}.
Furthermore, recent advances in recording technology focus on improving implant
durability and resolution~\cite{Steinmetz2021}, and on balancing the invasiveness
tradeoff by finding new, minimally invasive ways to record from closer to the cortex.
The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered through the
bloodstream via a catheter, removing the need for open brain surgery.
Nevertheless, surveys~\cite{Huggins2011, Huggins2015, Branco2021} in different
potential communication \ac{bci} user groups consistently show that non-invasive
technology is preferred over implanted electrodes, unless the added value of invasive
BCI is sufficiently large~\cite{Kageyama2020}.
This provides justification for focusing part of the current research effort on
non-invasive BCI.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical neurology
practice as well as in \ac{bci} research.
Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and offers
great temporal sampling resolution up to thousands of Hz.
\Ac{eeg} measures the electrical potential over time at a given electrode relative to a
reference electrode in volts (V), often on the scale of microvolts ($\mu$V).
As it is easily applicable on the outside of the scalp, many practical consumer-grade
\ac{eeg} headsets exist in the form of (often wireless) electrode caps or helmets.
Current consumer systems sometimes feature dry electrodes, but clinical and research
systems often use active electrodes with preamplifiers and electrolyte gel to reduce
impedance for improved signal quality, which slightly decreases practicality, since they
require a trained individual to properly administer.
Nevertheless, it is the most accessible \ac{bci} technology for both users and
researchers.

>>>>

Some impressive recent breakthroughs in speech and motor \ac{bci} for communication have
been realized using invasive microelectrode arrays~\cite{Willett2021} and
µ\Ac{ecog}~\cite{Metzger2023}.
Furthermore, recent advances in recording technology focus on improving implant
durability and resolution~\cite{Steinmetz2021}, and on balancing the invasiveness
tradeoff by finding new, minimally invasive ways to record from closer to the cortex.
The Synchron Stentrode~\cite{Mitchell2023}, for instance, can be delivered through the
bloodstream via a catheter, removing the need for open brain surgery.
Nevertheless, surveys~\cite{Huggins2011, Huggins2015, Branco2021} in different
potential communication \ac{bci} user groups consistently show that non-invasive
technology is preferred over implanted electrodes, unless the added value of invasive
BCI is sufficiently large~\cite{Kageyama2020}.
This provides justification for focusing part of the current research effort on
non-invasive BCI.

@@

\Ac{eeg} is the de facto non-invasive electrophysiology standard in clinical neurology
practice as well as in \ac{bci} research.
Developed since 1924~\cite{Berger1929}, it is cheap and relatively practical and offers
great temporal sampling resolution up to thousands of Hz.
\Ac{eeg} measures the electrical potential over time at a given electrode relative to a
reference electrode in volts (V), often on the scale of microvolts (µV).
As it is easily applicable on the outside of the scalp, many practical consumer-grade
\ac{eeg} headsets exist in the form of (often wireless) electrode caps or helmets.
Current consumer systems sometimes feature dry electrodes, but clinical and research
systems often use active electrodes with preamplifiers and electrolyte gel to reduce
impedance for improved signal quality, which slightly decreases practicality, since they
require a trained individual to properly administer.
Nevertheless, it is the most accessible \ac{bci} technology for both users and
researchers.

### 2024-12-04 chapters/bttda.tex
Proofread preface

\section{Experiments}
\subsection{Datasets and decoders}
We evaluated our proposed model in two offline EEG-based BCI decoding problems:
the event-related potential (ERP) and motor imagery (MI) paradigms using a
selection of the openly available MOABB benchmarking datasets~\cite{Aristimunha2023}.
Two ERP and two MI datasets were retained to reduce computational
demand.
Details about these datasets are found in \cref{tab:moabb}.
For the ERP datasets, the task is to distinguish target from non-target ERPs,
while the MI datasets consist of distinguishing different imagined limb
movements.
Within-session classification performance was assessed using stratified 5-fold
cross-validation to calculate the area under the receiver operating
characteristic curve (ROC-AUC).
\begin{table*}[t]
  \makebox[\textwidth][c]{
	  \input{chapters/bttda/moabb.tex}
  }
  \caption[MOABB datasets used for evaluation.]{MOABB datasets used for evaluation, with the number of
		subjects (Sub.), the number of EEG channels (Chan.), the number of
    classes (Cls.), the number of trials or trials per class for ERP
    datasets (Trials), the epoch length (Epoch len.), the sampling
		frequency (S. freq.), the number of sessions per subject (Sess.) and the
		number of runs.
    Adapted from \textcite{Aristimunha2023} and \textcite{Chevallier2024}.}
	\label{tab:moabb}
\end{table*}

@@

\Cref{tab:erp-score} lists the cross-validated ROC-AUC for all evaluated
decoders.
\begin{table}[t]
	\input{chapters/bttda/score_erp.tex}
  \caption[Within-session classification score for 2 event-related potential
  datasets.]{Area under the receiver operating characteristic curve for
		cross-validated within-session evaluation for HODA and our proposed decoders
		PARAFACDA and BTTDA evaluated on 2 event-related potential datasets.
    Scores for other decoders were taken from \textcite{Chevallier2024}.
		BTTDA reaches the highest performance for both evaluated datasets, closely
		followed by PARAFACDA.
	}
	\label{tab:erp-score}
\end{table}
The highest performance is achieved with the proposed BTTDA+LDA.
One-sided Wilcoxon signed-rank tests with a significance level of $\alpha=0.05$
reveal that both PARAFACDA+LDA and BTTDA+LDA significantly outperform
HODA+LDA in both the BNCI2014-008 (PARAFACDA+LDA: $p=0.0039$, BTTDA+LDA:
$p=0.0039$) and the BNCI2015-003 (PA\-RA\-FAC\-DA+LDA: $p=0.0001$, BTTDA+LDA:
$p=0.0049$) datasets.
No significant difference was found between BTTDA+LDA and PARAFACDA+LDA.

@@

\cref{tab:mi-score} lists the cross-validated classification scores for
the evaluated motor imagery datasets.
Note that, in line with~\textcite{Chevallier2024}, accuracy is reported for the
multi-class classification problem in BNCI2014-001, while ROC-AUC was reported for the
binary classification problem in BNCI2014-004.
\begin{table}[t]
	\input{chapters/bttda/score_mi.tex}
  \caption[Within-session classification score for 2 motor imagery datasets.]{Classification score for
		cross-validated within-session evaluation for HODA+LDA and our proposed decoders
		PARAFACDA+LDA and BTTDA+LDA evaluated on 2 motor imagery datasets.
		Scores for other decoders were taken from \textcite{Chevallier2024}.
    Accuracy is listed for BNCI2014-001, and \ac{rocauc} for BNCI2014-004.
    BTTDA outperforms HODA and PARAFACDA for BNCI2014-001 but does not reach
    performance comparable to the current state-of-the-art.
    For BNCI2014-004, PARAFACDA and BTTDA perform
    approximately on par with HODA and the current state-of-the art.}
	\label{tab:mi-score}
\end{table}
For the BNCI2014-001 dataset, HODA+LDA and our proposed decoders do not reach
satisfactory performance compared to the comparison methods, yet both
PARAFACDA+LDA and BTTDA+LDA improve upon HODA+LDA, and for BTTDA+LDA this
difference is significant ($p=0.0269$).
For the BNCI2014-004 dataset, the performance gap with comparison methods is
smaller, but no significant differences were found between HODA+LDA,
PARAFACDA+LDA, and BTTDA+LDA.

@@

Nevertheless, when considering the relative improvement over the original
HODA model, our results show that BTTDA+LDA performs consistently on par or
higher than HODA+LDA, with performances of respectively $86.43>83.25$ and
$85.08>82.57$ for \ac{erp} datasets and $55.13>53.51$ and $80.60<80.88$ for
\ac{mi} datasets.
The classification score obtained by BTTDA+LDA was also slightly
higher than that of the
restricted PARAFACDA+LDA for the ERP datasets ($86.43>86.19$ and $85.06>84.85$
respectively) and one of the MI datasets ($55.13>54.34$ and $80.60<80.88$
respectively), but these results were not significant, and further studies with
more datasets and subjects should show whether this holds.
\Cref{fig:blocks} shows that there is added value in finding extra blocks
over the first HODA block.
While no proof is given, we notice that NMSE monotonically decreases, indicating that
since the training score keeps increasing, this suggests that
eventually all the variation in the signal will be explained by the model
while still extracting features that are maximally discriminant.
Eventually, the number of blocks will reach a point of diminishing validation
score returns, when adding extra features to the decision classifier increases
its risk of overfitting instead of adding extra useful discriminatory
information.
Together with the improved classification scores presented, these results
point to the potential of our more flexible model block-term or
its special PARAFAC-structured case over a Tucker-structured model given proper block and
rank selection.
Since the optimal ranks for HODA+LDA were also determined through
cross-validation, BTTDA+LDA can improve over the first HODA block.
If validation shows that this is not possible, such as is the case for
BNCI2014-004, the BTTDA model is truncated to the first HODA block and little
performance is lost.

>>>>

\section{Experiments}
\subsection{Datasets and decoders}
We evaluated our proposed model in two offline EEG-based BCI decoding problems:
the event-related potential (ERP) and motor imagery (MI) pa\-ra\-digms using a
selection of the openly available MOABB benchmarking datasets~\cite{Aristimunha2023}.
Two ERP and two MI datasets were retained to reduce computational
demand.
Details about these datasets are found in \cref{tab:moabb}.
For the ERP datasets, the task is to distinguish target from non-target ERPs,
while the MI datasets consist of distinguishing different imagined limb
movements.
Within-session classification performance was assessed using stratified 5-fold
cross-validation to calculate the area under the receiver operating
characteristic curve (ROC-AUC).
\begin{table*}[t]
  \makebox[\textwidth][c]{
	  \input{chapters/bttda/moabb.tex}
  }
  \caption[MOABB datasets used for evaluation.]{MOABB datasets used for evaluation, with the number of
		subjects (Sub.), the number of EEG channels (Chan.), the number of
    classes (Cls.), the number of trials or trials per class for ERP
    datasets (Trials), the epoch length (Epoch len.), the sampling
		frequency (S. freq.), the number of sessions per subject (Sess.) and the
		number of runs.
    Adapted from \textcite{Aristimunha2023} and \textcite{Chevallier2024}.}
	\label{tab:moabb}
\end{table*}

@@

\Cref{tab:erp-score} lists the cross-validated ROC-AUC for all evaluated
decoders.
\begin{table}[t]
	\input{chapters/bttda/score_erp.tex}
  \caption[Within-session classification score for 2 \acs{erp}.]{Area under the receiver operating characteristic curve for
		cross-validated within-session evaluation for HODA and our proposed decoders
		PARAFACDA and BTTDA evaluated on 2 event-related potential datasets.
    Scores for other decoders were taken from \textcite{Chevallier2024}.
		BTTDA reaches the highest performance for both evaluated datasets, closely
		followed by PARAFACDA.
	}
	\label{tab:erp-score}
\end{table}
The highest performance is achieved with the proposed BTTDA+LDA.
One-sided Wilcoxon signed-rank tests with a significance level of $\alpha=0.05$
reveal that both PARAFACDA+LDA and BTTDA+LDA significantly outperform
HODA+LDA in both the BNCI2014-008 (PARAFACDA+LDA: $p=0.0039$, BTTDA+LDA:
$p=0.0039$) and the BNCI2015-003 (PA\-RA\-FAC\-DA+LDA: $p=0.0001$, BTTDA+LDA:
$p=0.0049$) datasets.
No significant difference was found between BTTDA+LDA and PA\-RA\-FAC\-DA+LDA.

@@

\cref{tab:mi-score} lists the cross-validated classification scores for
the evaluated motor imagery datasets.
Note that, in line with~\textcite{Chevallier2024}, accuracy is reported for the
multi-class classification problem in BNCI2014-001, while ROC-AUC was reported for the
binary classification problem in BNCI2014-004.
\begin{table}[t]
	\input{chapters/bttda/score_mi.tex}
  \caption[Within-session classification score for 2 \acs{mi} datasets.]{Classification score for
		cross-validated within-session evaluation for HODA+LDA and our proposed decoders
		PARAFACDA+LDA and BTTDA+LDA evaluated on 2 motor imagery datasets.
		Scores for other decoders were taken from \textcite{Chevallier2024}.
    Accuracy is listed for BNCI2014-001, and \ac{rocauc} for BNCI2014-004.
    BTTDA outperforms HODA and PARAFACDA for BNCI2014-001 but does not reach
    performance comparable to the current state-of-the-art.
    For BNCI2014-004, PARAFACDA and BTTDA perform
    approximately on par with HODA and the current state-of-the art.}
	\label{tab:mi-score}
\end{table}
For the BNCI2014-001 dataset, HODA+LDA and our proposed decoders do not reach
satisfactory performance compared to the comparison methods, yet both
PARAFACDA+LDA and BTTDA+LDA improve upon HODA+LDA, and for BTTDA+LDA this
difference is significant ($p=0.0269$).
For the BNCI2014-004 dataset, the performance gap with comparison methods is
smaller, but no significant differences were found between HODA+LDA,
PARAFACDA+LDA, and BTTDA+LDA.

@@

Nevertheless, when considering the relative improvement over the original
HODA model, our results show that BTTDA+LDA performs consistently on par or
higher than HODA+LDA, with performances of respectively $86.43>83.25$ and
$85.08>82.57$ for \ac{erp} datasets and $55.13>53.51$ and $80.60<80.88$ for
\ac{mi} datasets.
The classification score obtained by BTTDA+LDA was also slightly
higher than that of the
restricted PA\-RA\-FAC\-DA+LDA for the ERP datasets ($86.43>86.19$ and $85.06>84.85$
respectively) and one of the MI datasets ($55.13>54.34$ and $80.60<80.88$
respectively), but these results were not significant, and further studies with
more datasets and subjects should show whether this holds.
\Cref{fig:blocks} shows that there is added value in finding extra blocks
over the first HODA block.
While no proof is given, we notice that NMSE monotonically decreases, indicating that
since the training score keeps increasing, this suggests that
eventually all the variation in the signal will be explained by the model
while still extracting features that are maximally discriminant.
Eventually, the number of blocks will reach a point of diminishing validation
score returns, when adding extra features to the decision classifier increases
its risk of overfitting instead of adding extra useful discriminatory
information.
Together with the improved classification scores presented, these results
point to the potential of our more flexible model block-term or
its special PARAFAC-structured case over a Tucker-structured model given proper block and
rank selection.
Since the optimal ranks for HODA+LDA were also determined through
cross-validation, BTTDA+LDA can improve over the first HODA block.
If validation shows that this is not possible, such as is the case for
BNCI2014-004, the BTTDA model is truncated to the first HODA block and little
performance is lost.

### 2024-12-04 chapters/conclusion.tex
Proofread preface

However, the brain is a complex organ, so neural data is not linear in origin.
These regularizing assumptions based on prior knowledge are likely only
superficially true.
When optimizing for performance, it will soon become clear that, given a more
flexible model, there are
settings where these assumptions can be broken to better describe the ongoing
interactions.
For example, spatiotemporal beamforming or \ac{lda} with Kronecker-structured covariance
improves performance in low sample sizes but is outperformed by more flexible
methods, such as \ac{lda} using a Kronecker-sum covariance model or \ac{tlda}
when the training sample size increases.
Similarly, \ac{bttda} offers more flexibility.

This calls for the development of more flexible (multi-)linear models that still rely on some
structure but are combined with efficient model
selection, such as those proposed by \ac{bttda}.
These models try to combine the best of both worlds.

@@

The optimal approach should unify decoder development, choice of paradigm, and interface
design, as advocated by~\textcite{Pan2022} and~\textcite{Fouad2020}, by integrating both
engineering and clinical perspectives.
Performance and abilities likely depend to a great degree on the user,
so off-line decoding performance in healthy controls is probably not a good
predictor of on-line decoding performance and \ac{itr} in individuals with
\ac{sspi}.
Incremental gains in decoding performance for a specific paradigm might
also be outweighed by how well that paradigm is adapted to the user, factoring in
their skills and abilities.
Longitudinal research should follow several user groups, specifically those with gaze
impairment, and experiment with specific paradigms, interfaces, and decoders for
a given \ac{bci} user.

>>>>

However, the brain is a complex organ, so neural data is not linear in origin.
These regularizing assumptions based on prior knowledge are likely only
superficially true.
When optimizing for performance, it will soon become clear that, given a more
flexible model, there are
settings where these assumptions can be broken to better describe the ongoing
interactions.
For example, spatiotemporal beamforming or \ac{lda} with Kronecker-structured covariance
improves performance in low sample sizes but is outperformed by more flexible
methods, such as \ac{lda} using a Kronecker-sum covariance model or \ac{tlda}
when the training sample size increases.
Similarly, \ac{bttda} offers more flexibility than \ac{hoda}.

This calls for the development of more flexible (multi-)linear models that still rely on some
structure but are combined with efficient model
selection, such as the methods proposed in conjunction with \ac{bttda}.
These models try to combine the best of both worlds.

@@

The optimal approach should unify decoder development, choice of pa\-ra\-digm, and interface
design, as advocated by~\textcite{Pan2022} and~\textcite{Fouad2020}, by integrating both
engineering and clinical perspectives.
Performance and abilities likely depend to a great degree on the user,
so off-line decoding performance in healthy controls is probably not a good
predictor of on-line decoding performance and \ac{itr} in individuals with
\ac{sspi}.
Incremental gains in decoding performance for a specific paradigm might
also be outweighed by how well that paradigm is adapted to the user, factoring in
their skills and abilities.
Longitudinal research should follow several user groups, specifically those with gaze
impairment, and experiment with specific paradigms, interfaces, and decoders for
a given \ac{bci} user.

### 2024-12-04 chapters/cv.tex
Proofread preface

\subsection*{About me}
I am a researcher at KU Leuven Laboratory for Neuro- and Psychophysiology
and the University of Lille Research Center for Informatics, Signal Processing,
and Control Science. My research is centered on developing brain-computer
interface (BCI) technologies with a primary focus on assisting individuals with
severe disabilities, especially those with motor and eye movement impairments.
My goal is to enhance their quality of life by creating accurate and
user-friendly BCIs. Combining my expertise in machine learning, signal
processing, and user interface design with insights from neurophysiology, my
PhD centers around the development of an event-related potential BCI system
using EEG for real-time communication. I strongly believe in the potential of
innovative neurotechnology to empower individuals and patients, enabling them
to lead more independent and fulfilling lives.

\subsection*{Keywords}
brain-computer interfaces, event-related
potentials, (multi)linear classification, covert visual attention

@@

\begin{cvlist}
	\cvitem{PhD\ in Biomedical Sciences}{KU Leuven, 2024}
	\cvitem{PhD\ in Control Science and Signal Processing}{University of
		Lille, 2024}
	\cvitem{M.Sc.\ in Engineering Science: Computer Science}{KU Leuven, 2024}
	\cvitem{B.Sc.\ in Informatics}{KU Leuven, 2017}
\end{cvlist}

@@

\subsubsection*{Master theses and internships}
\begin{cvlist}
	\cvitem{Eye-tracker and ERP data fusion for gaze-independent visual
		BCI}{\ \\ Juliette  Meunier, University of Lille, 2024}
	\cvitem{LDA in	the combined
		space-time-frequency domain for BCI decoding}{\ \\Lunkyadi Sucipto, KU Leuven,
		2023--2023}
	\cvitem{Tackling the Midas Touch problem in eye tracking with a
		BCI}{\ \\Reniflal Ebenezer Sundaralal, KU Leuven, 2023--2023}
	\cvitem{An EEG connectivity analysis of Alzheimer’s Disease
		and Frontotemporal De\-men\-tia}{Zoe Barinaga, KU Leuven, 2023--2024}
	\cvitem{Single-trial ERP latencies as a predictor for the mode
		of visual attention}{\ \\Yildiz Dilara Parry, KU Leuven, 2023}
	\cvitem{A hybrid P300-gaze BCI alternative for navigating virtual
		spaces}{\ \\Gijs Claes, KU Leuven, 2021--2022}
\end{cvlist}

>>>>

%\subsection*{About me}
%I am a researcher at KU Leuven Laboratory for Neuro- and Psychophysiology
%and the University of Lille Research Center for Informatics, Signal Processing,
%and Control Science. My research is centered on developing brain-computer
%interface (BCI) technologies with a primary focus on assisting individuals with
%severe disabilities, especially those with motor and eye movement impairments.
%My goal is to enhance their quality of life by creating accurate and
%user-friendly BCIs. Combining my expertise in machine learning, signal
%processing, and user interface design with insights from neurophysiology, my
%PhD centers around the development of an event-related potential BCI system
%using EEG for real-time communication. I strongly believe in the potential of
%innovative neurotechnology to empower individuals and patients, enabling them
%to lead more independent and fulfilling lives.
%
%\subsection*{Keywords}
%brain-computer interfaces, event-related
%potentials, (multi)linear classification, covert visual attention

@@

\begin{cvlist}
	\cvitem{Ph.D.\ in Biomedical Sciences}{KU Leuven, 2024}
	\cvitem{Ph.D.\ in Control Science and Signal Processing}{University of
		Lille, 2024}
	\cvitem{M.Sc.\ in Engineering Science: Computer Science}{KU Leuven, 2024}
	\cvitem{B.Sc.\ in Informatics}{KU Leuven, 2017}
\end{cvlist}

@@

\subsubsection*{Master theses and internships}
\begin{cvlist}
	\cvitem{Eye-tracker and ERP data fusion for gaze-independent visual
		BCI}{\ \\ Juliette  Meunier, University of Lille, 2024}
	\cvitem{LDA in	the combined
		space-time-frequency domain for BCI decoding}{\ \\Lunkyadi Sucipto, KU Leuven,
		2023--2023}
	\cvitem{Tackling the Midas Touch problem in eye tracking with a
		BCI}{\ \\Reniflal Ebenezer Sundaralal, KU Leuven, 2023--2023}
	\cvitem{An EEG connectivity analysis of Dementias}{Zoe Barinaga, KU Leuven, 2023--2024}
	\cvitem{Single-trial ERP latencies as a predictor for the mode
		of visual attention}{\ \\Yildiz Dilara Parry, KU Leuven, 2023}
	\cvitem{A hybrid P300-gaze BCI alternative for navigating virtual
		spaces}{\ \\Gijs Claes, KU Leuven, 2021--2022}
\end{cvlist}

### 2024-12-04 chapters/patients.tex
Proofread preface

We evaluated the recorded data using the \ac{wcble}~\cite{VanDenKerchove2024}
and \ac{tlda}~\cite{Sosulski2022}
classifiers, as well as the Riemannian XDAWNCov+TS+LDA~\cite{Cecotti2017}.
For \ac{wcble}, a region of interest from 0 ms to 800 ms relative to stimulus
onset was used while the epoch was cropped to -100 ms to 900 ms. For the other
decoders, the epoch was cropped between 0 ms and 800 ms, which resulted in maximal
performance.
Decoding scores were obtained using 6-fold cross-validation where folds corresponded to
stimulation blocks.

>>>>

We evaluated the recorded data using the \ac{wcble}~\cite{VanDenKerchove2024}
and \ac{tlda}~\cite{Sosulski2022}
classifiers, as well as the Riemannian approach XDAWNCov+TS+LDA~\cite{Cecotti2017}.
For \ac{wcble}, a region of interest from 0 ms to 800 ms relative to stimulus
onset was used while the epoch was cropped to -100 ms to 900 ms. For the other
decoders, the epoch was cropped between 0 ms and 800 ms, which resulted in maximal
performance.
Decoding scores were obtained using 6-fold cross-validation where folds corresponded to
stimulation blocks.

### 2024-12-04 chapters/preface.tex
Proofread preface

\chapter{Preface}
The mind works in mysterious ways\ldots\
Let's try to hack it!

I'd like to acknowledge

>>>>

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

The past four and a half years have been a stimulating, instructive, and intense
journey of scientific and self-discovery. What started as a vague interest
in the applications of machine learning in the biomedical domain rapidly evolved
into a fascination with neural signal processing when I came into contact with
it via my master thesis.
Ultimately, this was supercharged by the very high science-fiction factor
of Brain-Computer Interfaces.
There is something about both the brain and computers, and
their beautiful complexity arising from simplicity, that I just can't
seem to let go of.

Yet, it is the tangible impact on people's lives of neurotechnology
which first drew me to the field.
The biggest advantage my PhD project had to offer was
its holistic, end-to-end nature, involving everything from innovation in signal
processing and machine learning under the hood to EEG
experimentation and, eventually, working with end-user patients. I have
always considered this PhD, first and foremost, a learning experience.
An experience that has provided me -- after a lot of trial and error -- with the
toolkit to, hopefully, continue making this impact in the future by
advancing biomedical technology.

For this, I am grateful to my thesis supervisors. The combination of the
network, resources, advice, and feedback you provided helped this project
succeed.
I would specifically like to thank prof. Marc Van Hulle for the many
opportunities you have given me, prof. François Cabestaing for your close,
active, and involved supervision, and prof. Hakim Si-Mohammed for the many
sharp-witted ideas and contributions you brought to the table.
I also want to thank the members of the jury, prof. Andrea Kübler, prof.
Fabien Lotte, prof. Reinhold Scherer, prof. Maarten De Vos, and prof.
Adalberto Simeone, for the time and effort you have spent evaluating
this thesis and for your insightful feedback.

Furthermore, I would like to explicitly thank everyone else who helped
in realizing the work presented in this thesis. Juliette, your
proactive attitude and multidisciplinary perspective were an immense help in
data collection. Without you, this thesis would probably be missing its
last chapter, and I hope you enjoy your well-deserved upcoming academic
career.
Working with some of our end-user study participants would also
not have been possible without the medical advice of prof. Etienne
Allart from the Neurorehabilitation Unit at CHU Lille, as well as
the others involved in recruiting and interacting with
these participants: Alixe and dr. Edward Schietekatte at TRAINM Neuro
Rehab Clinics, prof. Kristl Claeys and prof. Philip Van Damme from the
Neuromuscular Reference Center at UZ Leuven, Marie and the team at
Fondation Partage et Vie, and all colleagues who helped me in data
collection and processing. Most importantly, I am grateful to everyone
who participated in our study, and I am sincerely moved by the positive
reception among those of you who could benefit from it. Of course, I
would also like to thank all other students, colleagues, and friends who
submitted themselves to having their hair messed up by yet another EEG
experiment.

Since my thesis was a joint PhD between two labs, one at KU Leuven in
Belgium and one at the University of Lille in France, I got to
experience working at two different research institutions
This formed an invaluable enrichment of my career until now.
It also means I've had the amazing luck of having double the amount of wonderful
colleagues.

In Lille, I've very much appreciated working with Gaël. Your presence
never failed to lighten up the mood, and I wish you smooth sailings
in completing your own PhD. Next, I want to thank Jimmy for showing
me the way through Lille and its campus life, and for all the tea. I
also enjoyed discussing with José, Jean-Marc, and Marie-Hélène.
Even though I was not always physically
present in Lille, you've always made sure I felt part of the team.
Finally, I'm glad that the SIGMA team was around to keep me in touch with
my computer science roots.

In Leuven, I want to begin with a special thanks to those of you who
guided me at the start of my PhD: Benjamin, for introducing me to
the wonderful world of ERPs and linear decoding, and Axel and Arno for
powering through COVID with me. As for the rest of you lot -- Aline,
Anahit, Aurélie, Barbara, Bob, Elvira, Eva, Li Ang, Liuyin, Mani, Mansoureh,
Qiang, Tjaša, Valentina, and Yide -- you've made my days brighter in many ways,
and I could always count on you, be it for advice, support or help.
I also wish to thank the students and interns I was fortunate enough to
supervise for their dedication and great work, as well as the students at NeuroTech Leuven.
You never failed to impress me with creative ideas and excellent
execution. A final shoutout to everyone in and around the
Neurophysiology lab: Alessia, Astrid, Christophe, Inez, Jonathan,
Kenneth, Lara, Noa, Saba, Sara, Sofie, Stijn, Wouter, and all others.

On a more personal note, I probably wouldn't have gotten here if it wasn't
for the support (and proofreading) I have \emph{always} experienced
from my family and friends.
Cato, Loes, Mama, and Papa, I think you've provided about the
best nest I could imagine to grow up in. To Aderik, Joran and Siebe: thank you for
carrying me throughout my university studies.
To Hanne, Ingrid, Jonah, Joshua, Kiefer, Mathi, Vik, Ward and Yana, I have but one
thing to say: Snek \emph{is} love, Snek \emph{is} life.

Some say that whoever is mentioned first on a paper is the most
important, since they're supposed to have done all the work.
Without the last author on there, however, most scientific publications would
either never materialize or go unnoticed.
Therefore, I would like to mention my partner Liesel here at the end.
Liesel, thank you for everything you do and have done
for me. This does not only include the past few months and your help with
practical arrangements, proofreading, and putting up with me as I was
-- maybe a bit too much -- focused on work, but much, much more
than that.
I still cannot believe how lucky I am to have you by my side.

As you turn this page and start reading the first chapter, I will also
start a new chapter in my life. You, on the other hand, will dive into
some 200 pages of algebra, convoluted procedures, and tables with
numbers. Enjoy!

### 2024-12-04 chapters/wcble.tex
Proofread preface

Instead, we opt for a more robust, probabilistic definition of class latencies.
This robust estimation method yields latencies that
\begin{enumerate*}[label=(\arabic*)]
  \item are more meaningful as input for the second-stage
    classifier, and
  \item lead to smoother convergence in our proposed iterative alignment scheme
    for \ac{wcble}, which heavily relies on exact latency estimation.
\end{enumerate*}
Assume classifier $\mathcal{C}(\theta,f,\Pr)$ now can also output a probability
per class $\Pr(l|
\mat{X}[:,s_1:s_2)], \theta)$ for a given epoch $\mat{X}$, a feature of many
common classifiers.
Analogous to equation~\ref{eq:score}, we can now write
\begin{equation}
\Pr(\mat{X},\theta,l,s) = \frac{1}{R}\Pr(\mat{X}[:,s:s+(s_2-s_1)],\theta,l)
  \quad \forall s\in 1,\ldots,R
	\label{eq:prob}
\end{equation}
The latency features assuming the epoch belongs to a given class
$l\in\{\textrm{target}, \textrm{non-target}\}$ are then defined as the median of
the corresponding distributions
\begin{equation}
  s_l =\mathrm{median}\left[\Pr(s|\mat{X},\theta,l) \right]
  \label{eq:prob-latency}
\end{equation}
Note that $\Pr(s|\mat{X},\theta,\textrm{non-target}) =1-
  \Pr(s|\mat{X},\theta,\mathrm{target})$.
The median of the probability distribution over time is more robust to
outliers and noise than the maximum or minimum score.
For the non-target case, the median approach tends towards the center of a
near-uniform distribution, resulting in a more consistent latency estimate over
trials as compared to the minimum approach.

>>>>

Instead, we opt for a more robust, probabilistic definition of class latencies.
This robust estimation method yields latencies that
\begin{enumerate*}[label=(\arabic*)]
  \item are more meaningful as input for the second-stage
    classifier, and
  \item lead to smoother convergence in our proposed iterative alignment scheme
    for \ac{wcble}, which heavily relies on exact latency estimation.
\end{enumerate*}
Assume classifier $\mathcal{C}(\theta,f,\Pr)$ now can also output a probability
per class $\Pr(l|
\mat{X}[:,s_1:s_2)], \theta)$ for a given epoch $\mat{X}$, a feature of many
common classifiers.
Analogous to equation~\ref{eq:score}, we can now write
\begin{equation}
\Pr(\mat{X},\theta,l,s) = \frac{1}{R}\Pr(\mat{X}[:,s:s+(s_2-s_1)],\theta,l)
  \quad \forall s\in 1,\ldots,R
	\label{eq:prob}
\end{equation}
The latency features assuming the epoch belongs to a given class given by
$l\in\{\textrm{target}, \textrm{non-target}\}$ are then defined as the median of
the corresponding distributions
\begin{equation}
  s_l =\mathrm{median}\left[\Pr(s|\mat{X},\theta,l) \right]
  \label{eq:prob-latency}
\end{equation}
Note that $\Pr(s|\mat{X},\theta,\textrm{non-target}) =1-
  \Pr(s|\mat{X},\theta,\mathrm{target})$.
The median of the probability distribution over time is more robust to
outliers and noise than the maximum or minimum score.
For the non-target case, the median approach tends towards the center of a
near-uniform distribution, resulting in a more consistent latency estimate over
trials as compared to the minimum approach.

### 2024-12-04 cover_kul.tex
Proofread preface

  \vfill
  \begin{minipage}[b]{.6\textwidth}
    \raggedright
    Jury: \\
    Co-supervisor KU Leuven: \emph{prof. Marc Van Hulle} \\
    Co-supervisor Univ. Lille: \emph{prof. François Cabestaing} \\
    Chair examining committee: \emph{prof. John Creemers} \\
    Chair public defense: \emph{prof. Koen Poesen} \\
    \emph{prof. Andrea K\"ubler} \\
    \emph{prof.\ Fabien Lotte} \\
    \emph{prof.\ Reinhold Scherer} \\
    \emph{prof.\ Maarten De Vos} \\
    \emph{prof.\ Adalberto Simeone}
  \end{minipage}\hfill%
  \begin{minipage}[b]{.31\textwidth}
      \raggedright
      Dissertation presented in partial fulfillment of the requirements for
      the double degree of \emph{Doctor of Biomedical Sciences} (KU Leuven)
      and \emph{Doctor of Control Science and Signal Processing} (University
      of Lille)
  \end{minipage}
  \bigskip

>>>>

  \vfill
  \begin{minipage}[b]{.6\textwidth}
    \raggedright
    Jury: \\
    Co-supervisor KU Leuven: \emph{prof.\ Marc Van Hulle} \\
    Co-supervisor Univ. Lille: \emph{prof.\ François Cabestaing} \\
    Chair examining committee: \emph{prof.\ John Creemers} \\
    Chair public defense: \emph{prof.\ Koen Poesen} \\
    \emph{Prof.\ Andrea K\"ubler} \\
    \emph{Prof.\ Fabien Lotte} \\
    \emph{Prof.\ Reinhold Scherer} \\
    \emph{Prof.\ Maarten De Vos} \\
    \emph{Prof.\ Adalberto Simeone}
  \end{minipage}\hfill%
  \begin{minipage}[b]{.31\textwidth}
      \raggedright
      Dissertation presented in partial fulfillment of the requirements for
      the double degree of \emph{Doctor of Biomedical Sciences} (KU Leuven)
      and \emph{Doctor of Control Science and Signal Processing} (University
      of Lille)
  \end{minipage}
  \bigskip
